Proving Logical Atomicity using Lock Invariants\thanks{Supported by the National Science Foundation via grant \#1811894.}

### 2022-05-15 concurrency/ASL 2022/Verification of Concurrent Binary Search Tree.tex
edits throughout

%revise
For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. %describe weird binder
Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple
%Maybe say what a reasonable $Q$ is and why such a spec is still useful, specifically w.r.t. the situation where one of the other threads' action following the linearization point removes the element just inserted.

@@

Later, the TaDA logic~\cite{tada} presented an alternative set of lock specifications, based on the newly introduced notion of logical atomicity:
$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\ p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_insert}\ t\ k\ v)\}

\{\mathsf{BST}\ p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\ p\ t \land \mathrm{abs\_lookup}\ t\ k = v\}

\{\mathsf{BST}\ p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\ p\ (\mathrm{abs\_delete}\ t\ k)\}
\end{mathpar}
where $\mathsf{BST}\ p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$. For instance, $t$ may be a recursively defined tree type in Coq, and $\mathrm{abs\_insert}$ may be a pure Coq function that adds a key-value pair to such a tree; $\mathsf{BST}$ could then be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, and $p$ could be the location of the root node.

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
\begin{mathpar}
\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure. %a bit more about how we prove these, and what the next couple sections are doing about them

\subsection{Atomicity with lock invariants}%make clearer the connection to atomic specs
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. With the invariant-style lock specs, we can make $\mathsf{BST}\ p\ t$ the invariant of this lock: the lock then holds full ownership of the data structure, and provides it to whichever thread is currently executing an operation on the tree at $p$.
In this \emph{coarse-grained} locking implementation, any thread with partial ownership of the lock can acquire the lock, perform all the operations of the sequential insertion/lookup/deletion, and then release the lock. Naively, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock---but this is not enough to tell us that we have changed the lock's state by the end of the operation, since we merely restore the invariant that there is \emph{some} tree $t$ at $p$. This is where ghost state and atomic triples come in: we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$. %clarify share
Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then satisfy the atomic postcondition with one half of the ghost state and return the other half to the lock invariant. The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory. This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock. They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g. the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g. P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.

@@

Note that in this case, the private precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node. A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state. More generally, in this approach there is no need to track ownership of the locks of individual nodes, a source of considerable complexity in the lock-invariant-based specs. %This comes later in the current version.

@@

We demonstrate our approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_insert}\ m\ k\ v)\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land \mathrm{abs\_lookup}\ m\ k = v\rangle

\langle t.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{abs\_delete}\ m\ k)\rangle
\end{mathpar}
where insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

\ignore{\subsection{BST code}
\label{selflock}

@@

The atomic specification for \lstinline{delete} is:
$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (\mathrm{abs\_delete}\ m\ k)\rangle$$

@@

@@

\section{Conclusion}
\label{conclusion}
While TaDA-style atomic specs for locks have been widely adopted%?
, the implications of the switch away from invariant-based specs had not been thoroughly examined. We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations). The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures. When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs. Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof will be a considerable effort, using the newer style of lock specification will simplify all future proofs about lock-based data structures.

>>>>

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition has been formalized in terms of linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada} of the form $$\langle a.\ P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$ where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. This triple says that $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. The specific value of $a$ (the shared abstract state) can vary as other threads modify the data structure, as long as $P_p(a)$ is always maintained. For instance, the specification $$\langle s.\ \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, making no visible changes to whatever stack $s$ is in place until, at the linearization point, the current stack $s$ is replaced atomically with $v :: s$. The $\mathsf{stack}$ itself is treated as a publicly owned resource, and can only be accessed and modified atomically by threads holding the $\mathsf{is\_stack}$ assertion.

@@

Later, the TaDA logic~\cite{tada}, which codified the notion of logical atomicity, presented an alternative set of lock specifications:
$$\langle b.\ (\mathsf{L}(\ell) \land \neg b) \vee (\mathsf{U}(\ell) \land b) \rangle\ \texttt{acquire}(\ell)\ \langle \mathsf{L}(\ell) \land b \rangle \qquad \langle \mathsf{L}(\ell) \rangle\ \texttt{release}(\ell)\ \langle \mathsf{U}(\ell) \rangle$$
An \texttt{acquire} operation may access the lock repeatedly, finding it in either the locked ($\mathsf{L}$) or unlocked ($\mathsf{U}$) state; if it sees the unlocked state ($b$ holds), then it can atomically move to a locked state and return. A release simply moves atomically from the locked state to the unlocked state. These operations make no mention of a protected resource, but because they are atomic, they can interact with any global invariant (``shared region'' in TaDA's terminology): in particular, in combination with the invariant $(\mathsf{U}(\ell) * R) \vee \mathsf{L}(\ell)$, they can be used to derive the lock-invariant-based specs. It is also worth noting that these lock specifications are not axioms: they are proved for a simple spinlock implementation that uses atomic operations (e.g. compare-and-swap), which are the actual primitive operations in TaDA and the logics that follow it. In this style, tracking ownership of a lock is orthogonal to the lock assertion itself; a mechanism such as cancelable invariants~\cite{rustbelt-relaxed} can be used to both make the lock assertion $\mathsf{U}$/$\mathsf{L}$ publicly available and deallocate it once all threads are finished.

@@

Suppose we have a sequential implementation of a binary search tree. It has functions \texttt{insert}, \texttt{lookup}, and \texttt{delete}, proven to satisfy the following specifications:
\begin{mathpar}
\{\mathsf{BST}\:p\ t\}\ \texttt{insert}(p, k, v)\ \{\mathsf{BST}\:p\ (t[k \mapsto v])\}

\{\mathsf{BST}\:p\ t\}\ \texttt{lookup}(p, k)\ \{v.\ \mathsf{BST}\:p\ t \land t(k) = v\}

\{\mathsf{BST}\:p\ t\}\ \texttt{delete}(p, k)\ \{\mathsf{BST}\:p\ (t[k \mapsto \_])\}
\end{mathpar}
where $\mathsf{BST}\:p\ t$ asserts that some abstract binary search tree $t$ is represented in memory starting at location $p$. For instance, $t$ may be a recursively defined tree type in Coq, and $\mathsf{BST}$ would be defined as a recursive predicate that represents each node of $t$ as an object in memory, with fields for key, value, and pointers to left and right children, with the root node located at $p$.

A thread-safe version of this data structure should satisfy atomic specifications that closely correspond to the sequential specs:
\begin{mathpar}
\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto v])\rangle

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k \mapsto \_])\rangle
\end{mathpar}
These specifications assert that each operation appears to execute atomically, accessing and updating the state of the data structure without exposing any intermediate states---for instance, a \texttt{lookup} should not find a partially inserted key with a value other than the one passed to its \texttt{insert} call. We split the $\mathsf{BST}$ assertion into two pieces: $\nodeboxrep$ contains the points-to and lock-invariant assertions that a thread needs in order to call a BST function, and $\treerep$ contains the ghost state that describes the overall state of the tree and is treated as an atomically accessed shared resource. We change the abstract state of the data structure from a tree $t$ to a key-value map $m$ to reflect the fact that clients cannot observe the internal structure of the tree (which will be useful when we want to rearrange nodes during concurrent access). Our sequential implementation will almost certainly not satisfy these specifications, because its functions relied on being able to access any part of the data structure at any time: the $\mathsf{BST}$ assertion included full ownership of every location in the data structure.

To show that an implementation satisfies these atomic specifications, we must show that 1) the implementation can execute safely without owning any piece of the public assertion $\treerep$, only accessing it atomically and restoring it up until the linearization point, and 2) at some linearization point, the implementation atomically transforms the current map $m$ into one that satisfies the public postcondition. The linearization point may be a (physically or logically) atomic operation, or it may occur between steps of the program (e.g., if the current thread holds a lock on the modified section of the data structure, so other threads cannot distinguish between the original state and the modified state until the lock is released). In the following sections, we discuss the general structure of these proofs for lock-based implementations using each of the two kinds of lock specifications.

\subsection{Atomicity with lock invariants}
The easiest way to synchronize a data structure is to add a lock that each operation acquires at the start and releases at the end. These \emph{coarse-grained} locking implementations are easy to verify with invariant-style lock specs: the lock's invariant can contain the whole data structure assertion (e.g. $\mathsf{BST}\:p\ t$), so that whichever thread is currently executing an operation on the tree at $p$ has full ownership of the data structure. If we only want to prove that the operation is thread-safe, we might choose as our lock invariant $\exists t.\ \mathsf{BST}\ p\ t$, asserting that we gain full ownership of the tree in some state $t$ every time we acquire the lock. On the other hand, if we want to prove specifications that describe changes to the data structure---as in the atomic specification $\langle m.\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \treerep\ (m[k \mapsto v])\rangle$---we need ghost state to track the changes.

Specifically, we can define a piece of ghost state $\mathsf{ghost\_bst}\ t$ that is divided between the lock invariant and the public pre- and postconditions of the functions, connecting the tree in memory to the public abstract state of the data structure. For instance, we might choose a lock invariant $R_{\mathit{cg}} \triangleq \exists t.\ \mathsf{BST}\ p\ t * \mathsf{ghost\_bst}_{.5}\ t$, and then set $\nodeboxrep\ p \triangleq p \islock_\pi R_{\mathit{cg}}$ and $\treerep\ t \triangleq \mathsf{ghost\_bst}_{.5}\ t$. %clarify share
Then we can prove the atomic triples above, where at the linearization point we access $\treerep$, combine the two halves of the ghost state and update them to reflect the new state of the concrete data structure, and then fulfill the atomic postcondition with one half of the ghost state and return the other half to the lock invariant. The atomic postcondition reflects the fact that the abstract state has been accessed or changed, while the lock invariant maintains the connection between the abstract state and the actual data structure in memory. This approach to making changes to a locked data structure visible via ghost state is due to Zhang and Jung's proofs of an atomic syncer in Iris~\cite{atomic-syncer}.%, a higher-order function that wraps any arbitrary functionality in a lock. They proved that if a target function $f$ has a spec of the form $\{R * P\}\ f\ \{R * Q\}$, where $R$ is e.g. the ownership assertion for a data structure, then the locked version $f'$ of $f$ has spec $\langle g. P * \mathsf{ghost}_{.5}\ g\rangle\ f'\ \langle Q * \mathsf{ghost}_{.5}\ g'\rangle$.

@@

Note that in this case, the private precondition $\mathsf{inFP}(n)$ is merely the knowledge that $n$ is in the data structure, and does not involve any ownership---in particular, a caller does not need a share of the lock for the root node. A thread can acquire the lock for any node $n$ in the data structure by atomically accessing $\mathsf{CSS}$, extracting the sub-assertion corresponding to $n$, and switching it to the locked state. More generally, in this approach there is no need to track ownership of the locks of individual nodes, which as we will see in section~\ref{selflock} is a source of considerable complexity in lock-invariant-based proofs.

@@

We demonstrate our approach to proving logically atomic specifications with invariant-based lock specs by verifying a binary search tree (BST) with fine-grained locking in VST, using the approach of section~\ref{atomicity} to prove atomic specifications for the BST operations while using invariant-style specs for our locks (as required by VST's soundness proof).
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto v])\rangle

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ m \land m(k) = v\rangle

\langle m.\ \nodeboxrep\ p\ |\ \treerep\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (m[k\mapsto \_])\rangle
\end{mathpar}
where insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract key-value map $m$ implemented by the tree. In this section, we describe the construction of these predicates, and the proofs of the operations, for a BST implementation based on hand-over-hand locking, using lock-invariant-based specs for our locks.

@@

\ignore{\subsection{BST code}

@@

The atomic specification for \lstinline{delete} is:
$$\langle m.\ \nodeboxrep_g\ p\ |\ \treerep_g\ m\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep_g\ p\ |\ \treerep_g\ (m[k \mapsto \_])\rangle$$

@@

VeriFast~\cite{verifast}, a separation-logic-based verifier for C and Java programs, also supports lock-based and atomic concurrency~\cite{verifast-conc}, and has been used to verify a hand-over-hand-locking linked list similar to that of Krishna et al. The specifications of both the lock operations and the list itself use a precursor of TaDA's logical atomicity. VeriFast is not a foundationally verified system, but its basic logic is verified against the semantics of a toy language in Coq.

@@

\section{Conclusion}
\label{conclusion}
While most new CSL proofs of lock-based data structures use TaDA-style atomic specs for their locks, the implications of the switch away from invariant-based specs had not been thoroughly examined. We have now demonstrated that invariant-based specs can still be used to prove logically atomic specs for larger data structures (even for fine-grained implementations) and that this requires somewhat more complex proofs than using atomic lock specs (especially for fine-grained implementations). The old style of lock spec may still be used in systems like VST, where switching would invalidate existing automation or soundness proofs, and in these systems the approach we have outlined can be used to obtain strong correctness properties for fine-grained-locking data structures. When possible, however, we would prefer to use atomic specifications for locks to simplify our proofs. Indeed, we intend to investigate shifting the foundations of VST to use atomic operations and derive atomic specs for locks: while updating the soundness proof will be a considerable effort, using the newer style of lock specification will simplify all future proofs about lock-based data structures.

### 2022-05-15 concurrency/ASL 2022/intro.tex
edits throughout

% Contributions:
% verified C implementation of BST with atomic specs -- the first? VeriFast did it, but not atomic specs
% technique for connecting locks with abstract state/atomicity
	% note that the flow interfaces team sort of does atomicity for locks, but they assume an atomic spec for locks that's also application-specific? no, they implement them in the TaDA style and prove an atomic spec.
% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract -- this corresponds to flow interfaces/CSS templates, basically
% delete proofs -- interesting interplay between non-visible range increase and lock-protected range modification
% lock-free implementation?

@@

Concurrent separation logic (CSL)~\cite{csl} is a useful tool for proving correctness of concurrent programs. While early iterations focused on data structures implemented with locks~\cite{gotsman,oraclesematic} and specifications involving memory safety and data-race-freedom, more recent logics are able to prove linearizability-style functional correctness~\cite{tada} of lock-free implementations that use low-level atomic memory operations~\cite{iris}, even under weak memory models~\cite{gps}%update cite
. In particular, \emph{logical atomicity}~\cite{tada} has become the gold standard for concurrent data structure specifications, capturing the idea that data structure operations should appear to take place instantaneously with no externally visible intermediate states. The effects of a logically atomic operation become visible at a \emph{linearization point}, a single instruction (usually an atomic memory operation or call to a logically atomic function) that publishes the new state of the data structure.

Intuitively, a critical section in a lock-based implementation serves the same role as an atomic instruction in a lock-free data structure, and it is not hard to identify linearization points in lock-based implementations---the effects of an update become visible when the lock protecting the updated data is released. The translation of this intuition into proof, however, depends on the specifications used for lock operations. The CSL literature contains two distinct specifications for lock acquire and release:
\begin{enumerate}
\item In Gotsman/Hobor-style specs, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
\item In TaDA-style specs, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
\end{enumerate}
Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}. However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.

In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t. the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs. We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}. The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access. However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise. To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique may be useful in tools like VST where the older style is deeply integrated into the tool.

>>>>

% Contributions:
% verified C implementation of BST with atomic specs -- the first? VeriFast did it, with a precursor to atomic specs
% technique for connecting locks with abstract state/atomicity
	% note that the flow interfaces team sort of does atomicity for locks, but they assume an atomic spec for locks that's also application-specific? no, they implement them in the TaDA style and prove an atomic spec.
% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract -- this corresponds to flow interfaces/CSS templates, basically
% delete proofs -- interesting interplay between non-visible range increase and lock-protected range modification
% lock-free implementation?

@@

Concurrent separation logic (CSL)~\cite{csl} is a useful tool for proving correctness of concurrent programs. While early iterations focused on data structures implemented with locks and specifications involving memory safety and data-race-freedom~\cite{gotsman,oraclesematic}, more recent logics are able to prove linearizability-style functional correctness of lock-free implementations that use low-level atomic memory operations~\cite{tada,iris}, even under weak memory models~\cite{rustbelt-relaxed}. In particular, \emph{logical atomicity}~\cite{tada} has become the gold standard for concurrent data structure specifications, capturing the idea that data structure operations should appear to take place instantaneously with no externally visible intermediate states. The effects of a logically atomic operation become visible at a \emph{linearization point}, a single instruction (usually an atomic memory operation or call to a logically atomic function) that publishes the new state of the data structure.

Intuitively, a critical section in a lock-based implementation serves the same role as an atomic instruction in a lock-free data structure, and it is not hard to identify linearization points in lock-based implementations---the effects of an update become visible when the lock protecting the updated data is released. The translation of this intuition into proof, however, depends on the specifications used for lock operations. The CSL literature contains two distinct specifications for lock acquire and release:
\begin{enumerate}
\item In Gotsman/Hobor-style specs~\cite{gotsman,oraclesematic}, each lock is associated with a lock invariant $R$, which is gained by \texttt{acquire} operations and restored by \texttt{release} operations.
\item In TaDA-style specs~\cite{tada}, \texttt{acquire} atomically moves a lock from the unlocked state to the locked state, and \texttt{release} does the reverse.
\end{enumerate}
Style 1 can be derived from style 2, and is also a natural fit for atomicity proofs of larger data structures, so it has been the style of choice for most recent verifications~\cite{tada-live,templates}. However, unpublished work by Zhang~\cite{atomic-syncer} shows that style 1 can also be used to prove atomic specifications for data structures.

In this paper, we adapt this approach to the Verified Software Toolchain~\cite{plfcc}, a system for proving correctness of C programs that relies on style 1 lock specs for its soundness proof w.r.t. the CompCert compiler~\cite{compcert}, and has recently been extended with support for general ghost state and atomicity proofs. We demonstrate that 1) \textbf{our approach can be used to prove atomic specs for interesting data structures with fine-grained locking}, and 2) \textbf{these proofs require some additional complications that can be avoided with style 2 lock specs}. The key difference is that in style 1 a thread must gain ownership of the lock invariant assertion for a component before acquiring the lock and reading/modifying the component, while in style 2 the locks and components alike can be considered part of a single abstract state that is accessed at each lock access. However, the top-level specifications proved for the data structure are not affected by the complexity of the lock specs, so proofs of clients of the data structure are no more complex than they would be otherwise. To our knowledge, this is the first formal comparison of the two styles of lock specifications, and our technique should be useful in tools like VST where the older style is deeply integrated into the tool.
